---
title: "Preparing Climate Data for Use With LANDIS-II"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing Climate Data for Use With LANDIS-II}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{arrow,BioSIM,climateR,climr,digest,elevatr,furrr,sf,terra,withr,zonal}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
VignetteDepends <- c(
  ## package list should match %\VignetteDepends above!
  "arrow",
  "BioSIM",
  "climateR",
  "climr",
  "digest",
  "elevatr",
  "furrr",
  "sf",
  "terra",
  "withr",
  "zonal"
)
all_pkgs_avail <- vapply(
  VignetteDepends,
  function(pkg) {
    require(pkg, character.only = TRUE, quietly = TRUE)
  },
  logical(1)
) |>
  all()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE, # all_pkgs_avail
  purl = FALSE
)
```

```{r setup}
library(landisutils)
```

## Define a 'study area' or 'area of interest'

```{r, study-area}
## random study area in BC using BEC zones as ecoregions
ecoregionPolys <- landisutils::test_ecoregionPolys

plot(ecoregionPolys["PolyID"])
```

## Historical weather data

```{r setup-climate}
clim_years <- 2018:2019 ## availability is 1980 to last year

clim_data_path <- withr::local_tempdir("climate_data_")
```

### Daily Climate Data (BioSIM)

Daily weather is fetched from BioSIM using the `ClimaticEx_Daily` model.
The high-level wrapper [`prep_daily_weather()`](../reference/prep_climate_data.html)
takes a study-area polygons object and a column name identifying ecoregions:

```{r, get-daily-weather}
climvars_daily <- c("prcp", "tmax", "tmin")

daily_weather <- prep_daily_weather(
  vars = climvars_daily,
  years = clim_years,
  studyArea = ecoregionPolys,
  id = "PolyID"
)

head(daily_weather)

clim_file <- file.path(clim_data_path, "climate-data-daily.csv")
writeClimateData(daily_weather, clim_file)
```

Internally, `prep_daily_weather()` chains four lower-level helpers that you can
call directly when you need finer control (e.g. to schedule per-batch fetches
as dynamic targets):

```{r, get-daily-weather-lowlevel, eval = FALSE}
## 1. Build an elevation raster covering the study area (lon-lat).
elev <- get_elevation_rast(ecoregionPolys, z = 9)

## 2. Tile cells of `elev` that fall inside `ecoregionPolys` into batches.
##    Each row carries Longitude/Latitude/Elevation, plus EcoID (from the
##    studyArea field named by `id`) and BatchID (for parallel dispatch).
locations_batch <- create_locations_df(
  elev = elev,
  studyArea = ecoregionPolys,
  id = "PolyID",
  batch_size = 500
)

## 3. Fetch one (batch, year) at a time. Cap parallelism to <= 4 workers
##    so we don't overload the BioSIM web service. The cache is namespaced
##    by a short hash of `studyArea` so distinct study areas don't collide.
future::plan(future::multisession, workers = 2)

sa_hash <- landisutils:::.studyArea_hash(ecoregionPolys)

grid <- expand.grid(batch = seq_along(locations_batch), year = clim_years)
furrr::future_map2(grid$batch, grid$year, function(b, y) {
  get_clim_daily(
    locations_batch[b], year = y,
    studyArea_hash = sa_hash, path = clim_data_path
  )
})

## 4. Read the partitioned Arrow CSV dataset, summarize by ecoregion,
##    and pivot to LANDIS-II Climate Library wide format.
daily_weather <- assemble_climate_library_file(
  dataset_path = file.path(clim_data_path, "ClimaticEx_Daily", sa_hash),
  vars = c("Prcp", "Tmin", "Tmax"),
  id_col = "EcoID"
)
```

#### Daily climate inputs for Social Climate Fire

The CSV that `prep_daily_weather()` / `assemble_climate_library_file()`
produces is consumed by the LANDIS-II Climate Library
(`Daily_RandomYears` / `Daily_AverageAllYears` time series). The
[Social Climate Fire](https://github.com/LANDIS-II-Foundation/Extension-Social-Climate-Fire)
extension reads daily fire-weather indirectly through the Climate Library and
computes FWI internally.
`assemble_climate_library_file_scf()` is a thin wrapper that emits the same
table with the lowercase `Variable` spellings used by the Social-Climate-Fire
reference input
([`LTB_ClimateInputs_91_10_v2.csv`](https://github.com/LANDIS-II-Foundation/Extension-Social-Climate-Fire/blob/master/Testing/Core8-SocialClimateFire4.0/LTB_ClimateInputs_91_10_v2.csv)):
`precip`, `maxtemp`, `mintemp`, `windspeed`, `winddirection`. The Climate
Library parser is case-insensitive, so either spelling works as input — pick
whichever your scenario template expects.

```{r, get-daily-weather-scf, eval = FALSE}
## populate the BioSIM cache with the 5 SCF variables
prep_daily_weather(
  vars = c("prcp", "tmax", "tmin", "ws", "wnddir"),
  years = clim_years,
  studyArea = ecoregionPolys,
  id = "PolyID"
)

sa_hash <- landisutils:::.studyArea_hash(ecoregionPolys)
scenario_tag <- landisutils:::.biosim_scenario_tag("RCP45", "RCM4")

scf_inputs <- assemble_climate_library_file_scf(
  dataset_path = file.path(
    clim_data_path, "ClimaticEx_Daily", sa_hash, scenario_tag
  ),
  vars = c("Prcp", "Tmin", "Tmax", "WndS", "WndD"),
  id_col = "EcoID"
)

clim_file <- file.path(clim_data_path, "scf-climate-inputs.csv")
writeClimateData(scf_inputs, clim_file)
```

### Monthly Climate Data (TerraClim)

As above, we can get monthly data using the following recipe:

```{r, get-terraclim-data}
climvars_monthly <- c("ppt", "tmax", "tmin")

monthly_weather <- prep_monthly_weather(
  vars = climvars_monthly,
  years = clim_years,
  studyArea = ecoregionPolys,
  id = "PolyID"
)

head(monthly_weather)

clim_file <- file.path(clim_data_path, "climate-data-monthly.csv")
writeClimateData(monthly_weather, clim_file)
```

We can also use this recipe to get monthly `AET` data to prepare the `EcoregionParameters` table:

```{r, get-terraclim-aet}
aet_df <- prep_monthly_weather(
  vars = "aet",
  years = clim_years,
  studyArea = ecoregionPolys,
  id = "PolyID"
)

head(aet_df)

erp_df <- prepEcoregionParameters(aet_df)

head(erp_df)
```

### Monthly Climate Data (BioSIM)

If you want monthly weather co-located with the daily BioSIM cells (same
elevation raster, same point batches), use `prep_monthly_weather_biosim()` —
it shares the elevation/batching machinery with `prep_daily_weather()` and
fetches BioSIM's `Climatic_Monthly` model. Wind variables (`ws`, `wnddir`)
are pulled in the same call from BioSIM's companion `ClimaticWind_Monthly`
model; wind from-direction is summarised across the model's 36 directional
bins (`W0`..`W350`) using a frequency-weighted circular mean:

```{r, get-biosim-monthly}
climvars_monthly_biosim <- c("prcp", "tmax", "tmin", "temp", "rh", "srad", "ws", "wnddir")

monthly_weather_biosim <- prep_monthly_weather_biosim(
  vars = climvars_monthly_biosim,
  years = clim_years,
  studyArea = ecoregionPolys,
  id = "PolyID"
)

head(monthly_weather_biosim)

clim_file <- file.path(clim_data_path, "climate-data-monthly-biosim.csv")
writeClimateData(monthly_weather_biosim, clim_file)
```

The lower-level helpers ([`get_clim_monthly()`](../reference/get_clim_monthly.html),
[`assemble_climate_library_file_monthly()`](../reference/assemble_climate_library_file_monthly.html))
mirror the daily pipeline if you need to schedule per-batch fetches as
dynamic targets:

```{r, get-biosim-monthly-lowlevel, eval = FALSE}
elev <- get_elevation_rast(ecoregionPolys, z = 9)
locations_batch <- create_locations_df(
  elev = elev,
  studyArea = ecoregionPolys,
  id = "PolyID",
  batch_size = 500
)

future::plan(future::multisession, workers = 2)

sa_hash <- landisutils:::.studyArea_hash(ecoregionPolys)

grid <- expand.grid(batch = seq_along(locations_batch), year = clim_years)
furrr::future_map2(grid$batch, grid$year, function(b, y) {
  get_clim_monthly(
    locations_batch[b], year = y,
    studyArea_hash = sa_hash, path = clim_data_path
  )
})

monthly_weather_biosim <- assemble_climate_library_file_monthly(
  dataset_path = file.path(clim_data_path, "Climatic_Monthly", sa_hash),
  vars = c("TotalPrcp", "MeanTmin", "MeanTmax", "WndS", "WndD"),
  id_col = "EcoID"
)
```

### Monthly Climate Data (climr)

[climr](https://github.com/bcgov/climr) provides ClimateNA-style change-factor
downscaling against a PRISM/DAYMET composite reference grid for North America.
Historical observational years span 1901-2024
(`obs_ts_dataset = "climatena"` or `"cru.gpcc"`). Wind variables are not
available from climr - use `prep_monthly_weather_biosim()` for wind.

```{r, get-climr-monthly}
climvars_monthly_climr <- c("prcp", "tmax", "tmin", "temp", "rh", "srad", "cmd")

monthly_weather_climr <- prep_monthly_weather_climr(
  vars = climvars_monthly_climr,
  years = clim_years,
  studyArea = ecoregionPolys,
  id = "PolyID"
)

head(monthly_weather_climr)

clim_file <- file.path(clim_data_path, "climate-data-monthly-climr.csv")
writeClimateData(monthly_weather_climr, clim_file)
```

The lower-level helpers ([`get_clim_monthly_climr()`](../reference/get_clim_monthly_climr.html),
[`assemble_climate_library_file_monthly_climr()`](../reference/assemble_climate_library_file_monthly_climr.html))
mirror the BioSIM pipeline if you need to schedule per-year fetches as
dynamic targets. climr's own on-disk cache is configured via
`options("climr.cache.path" = ...)`; `prep_monthly_weather_climr()` defaults
this to a `climr/` subdirectory of `getOption("landisutils.cache.path")` for
the duration of the call.

## Climate projections

The same `prep_*()` functions used above for historical/observational
weather can also produce **future** weather under climate-change scenarios.
Two complementary sources are wired in:

- **BioSIM** (CMIP5-era): `prep_daily_weather()` and
  `prep_monthly_weather_biosim()` accept `rcp` and `clim_model` arguments
  forwarded to `BioSIM::generateWeather()`. `rcp` is one of
  `"CONSTANT_CLIMATE"`, `"RCP45"`, `"RCP85"`; `clim_model` is one of
  `"RCM4"` (Canadian Regional Climate Model 4, CanESM2-driven), `"GCM4"`
  (CGCM4 / CanESM2-family GCM), or `"Hadley"` (HadGEM).
- **climr** (CMIP6): `prep_monthly_weather_climr()` accepts `gcms`, `ssps`,
  and `max_run` arguments forwarded to `climr::downscale()`. Pass any subset
  of `climr::list_gcms()` (or use the bcgov-recommended 8-member subset
  exposed as `climr_ensemble_8`) and any subset of `climr::list_ssps()`.

The BioSIM and climr caches are namespaced by scenario, so historical and
future fetches for the same study area coexist on disk without colliding.

```{r setup-climate-future}
clim_years_future <- 2041:2042
```

### Daily future weather (BioSIM, CanESM2 + RCP 4.5)

Pass `rcp` and `clim_model` straight through to `prep_daily_weather()`. The
Canadian fourth-generation GCM (CGCM4 / CanESM2-family) corresponds to
BioSIM's `clim_model = "GCM4"`:

```{r, get-daily-weather-future}
climvars_daily <- c("prcp", "tmax", "tmin")

daily_weather_future <- prep_daily_weather(
  vars = climvars_daily,
  years = clim_years_future,
  studyArea = ecoregionPolys,
  id = "PolyID",
  rcp = "RCP45",
  clim_model = "GCM4"
)

head(daily_weather_future)

clim_file <- file.path(clim_data_path, "climate-data-daily-rcp45-canesm.csv")
writeClimateData(daily_weather_future, clim_file)
```

The same arguments are also accepted by `prep_monthly_weather_biosim()`,
which lets you pull wind variables (`ws`, `wnddir`) under the matching
scenario - climr does not expose wind, so pair it with BioSIM when you need
wind under a future projection.

### Monthly future weather (climr, 8-member ensemble + SSP2-4.5)

`prep_monthly_weather_climr()` switches into projection mode whenever `gcms`
is non-`NULL`. The bcgov-recommended 8-member subset of
`climr::list_gcms()` is exposed for convenience as `climr_ensemble_8`; see
<https://bcgov.github.io/climr/articles/guidance_ensembleSelection.html>
for the rationale. SSP2-4.5 (`"ssp245"`) is the standard CMIP6 analogue to
CMIP5 RCP 4.5:

```{r, get-climr-monthly-future}
climvars_monthly_climr <- c("prcp", "tmax", "tmin", "temp", "rh", "srad", "cmd")

monthly_weather_climr_future <- prep_monthly_weather_climr(
  vars = climvars_monthly_climr,
  years = clim_years_future,
  studyArea = ecoregionPolys,
  id = "PolyID",
  gcms = climr_ensemble_8,
  ssps = "ssp245",
  max_run = 0L
)

head(monthly_weather_climr_future)

clim_file <- file.path(clim_data_path, "climate-data-monthly-climr-ssp245-ens8.csv")
writeClimateData(monthly_weather_climr_future, clim_file)
```

Each row is the **multi-GCM ensemble mean** across the 8 members, taken
inside `assemble_climate_library_file_monthly_climr()`, so the output has
the same `(Year, Month, Variable, <ecoregion-id>...)` shape as the
observational pipeline. Set `max_run > 0` to additionally average over
multiple runs per GCM (climr defaults to ensemble-mean only, `max_run = 0`).

<!-- cleanup -->

```{r cleanup, include = FALSE}
withr::deferred_run()
```
