---
title: "Preparing Dynamic Fire and Fuels Inputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing Dynamic Fire and Fuels Inputs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(landisutils)

tmp_pth <- withr::local_tempdir("example_DynamicFireFuels_")
```

## Sample extension inputs

### Dynamic Fire

All sample values from Dynamic Fire v4.0 test files (for LANDIS-II v8):
<https://github.com/LANDIS-II-Foundation/Extension-Dynamic-Fire-System/blob/master/testings/Core8-DynamicFire4.0/dynamic-fire_SetUpFire.txt>

```{r dynamic-fire-inputs}
fire_sizes <- tibble::tribble(
  ~EcoCode , ~EcoName , ~Mu , ~Sigma , ~Max ,
         1 , "fire1"  , 4.0 , 0.1    ,   50 ,
         2 , "fire2"  , 4.5 , 0.3    ,   50 ,
         3 , "fire3"  , 5.0 , 0.5    ,   50 ,
  ) |> dplyr::bind_cols(tibble::tribble(
    ~SpFMCLo , ~SpFMCHi , ~SpHiProp , ~SumFMCLo , ~SumFMCHi , ~SumHiProp , 
          85 ,      100 ,      0.50 ,        92 ,       120 ,       0.50 ,
          85 ,      105 ,      0.70 ,        94 ,       120 ,       0.40 ,
          85 ,      110 ,      0.60 ,        90 ,       120 ,       0.70
    )) |> dplyr::bind_cols(tibble::tribble(
     ~FallFMCLo , ~FallFMCHi , ~FallHiProp , ~OpenFuelIndex , ~NumFires ,
            120 ,        120 ,        0.50 ,              2 , 0.5       ,
            120 ,        120 ,        0.50 ,              2 , 0.2       ,
            120 ,        120 ,        0.50 ,             16 , 0.1
     ))

dyn_ecoregion <- prepDynamicEcoregionTable() ## optional

season <- tibble::tribble(
  ~Name    , ~LeafStatus , ~PropFire , ~PercentCuring , ~DayLengthProp ,
  "Spring" , "LeafOff"   , 0.20      ,              50 , 1.0            ,
  "Summer" , "LeafOn"    , 0.50      ,              51 , 1.0            ,
  "Fall"   , "LeafOff"   , 0.30      ,             100 , 1.0
)

fuel_type <- defaultFuelTypeTable()

fire_damage <- defaultFireDamageTable()

log_file <- file.path(tmp_pth, "fire/dynamic-fire-event-log.csv")
sum_log_file <- file.path(tmp_pth, "fire/summary-log.csv")

## these files don't need to be functional, just need to exist
ifrm_file <- file.path(tmp_pth, "Ecoregions.tif")
gs_file <- file.path(tmp_pth, "GroundSlope.tif")
ua_file <- file.path(tmp_pth, "UphillSlope.tif")
iwdb_file <- file.path(tmp_pth, "dynamic-fire_WeatherData.csv")

all_files <- c(ifrm_file, gs_file, ua_file, iwdb_file)
purrr::walk2(.x = rep("", length(all_files)), .y = all_files, .f = writeLines)
```

### Dynamic Fuels

All sample values from Dynamic Fuels v4.0 test files (for LANDIS-II v8):
<https://github.com/LANDIS-II-Foundation/Extension-Dynamic-Fire-System/blob/master/testings/Core8-DynamicFire4.0/dynamic-fire_SetUpFuel.txt>

```{r dynamic-fuels-inputs}
spp_fuel_coeffs <- tibble::tribble(
  ~Species   , ~FuelCoefficient ,
  "abiebals" , 1.00             ,
  "acerrubr" , 0.50             ,
  "acersacc" , 1.00             ,
  "betualle" , 1.00             ,
  "fraxamer" , 1.00             ,
  "piceglau" , 1.00             ,
  "pinubank" , 1.00             ,
  "pinuresi" , 1.00             ,
  "pinustro" , 1.00             ,
  "poputrem" , 1.00             ,
  "querelli" , 1.00             ,
  "querrubr" , 1.00             ,
  "thujocci" , 1.00             ,
  "tiliamer" , 1.00
)

## NOTE: `AgeMin` and `AgeMax` specified here (will be converted to range internally)
fuel_types <- tibble::tribble(
  ~FuelType , ~BaseFuel           , ~AgeMin , ~AgeMax , ~Species                                                                                                         ,
          1 , "Conifer"           ,       0 ,     900 , list("thujocci")                                                                                                 ,
          2 , "Conifer"           ,       0 ,     500 , list("piceglau", "abiebals")                                                                                     ,
          3 , "Conifer"           ,      41 ,     100 , list("pinubank")                                                                                                 ,
          4 , "Conifer"           ,       0 ,      40 , list("pinubank")                                                                                                 ,
         16 , "Open"              ,       0 ,      20 , list("pinustro", "pinuresi")                                                                                     ,
          5 , "Conifer"           ,      20 ,     400 , list("pinustro", "pinuresi")                                                                                     ,
          6 , "ConiferPlantation" ,       0 ,     100 , list("piceglau", "-abiebals", "-pinubank", "-pinustro")                                                          ,
          8 , "Deciduous"         ,       0 ,     300 , list("acerrubr", "acersacc", "betualle", "fraxamer", "poputrem", "betupapy", "querelli", "querrubr", "tiliamer")
)

disturb_conv <- tibble::tribble(
  ~Fuel , ~Type , ~Duration       , ~Prescription    ,
     14 ,    20 , "WindSeverity3" , "AspenClearcut"  ,
     13 ,    20 , "WindSeverity4" , "MaxAgeClearcut" ,
     15 ,    20 , "WindSeverity5" , ""
)
```

## Extension configuration

### Dynamic Fire

```{r dynamic-fire-write}
## create the dynamic fire extension config object
ext_dyn_fire <- DynamicFire$new(
  path = tmp_pth,
  Timestep = 10,
  EventSizeType = "size_based",
  BuildUpIndex = "yes",
  WeatherRandomizer = 0L,
  FireSizesTable = fire_sizes,
  InitialFireEcoregionsMap = ifrm_file,
  DynamicEcoregionTable = dyn_ecoregion, ## optional
  GroundSlopeFile = gs_file,
  UphillSlopeAzimuthMap = ua_file,
  SeasonTable = season,
  InitialWeatherDatabase = iwdb_file,
  DynamicWeatherTable = NULL, ## not used
  FuelTypeTable = fuel_type,
  SeverityCalibrationFactor = 1.0,
  FireDamageTable = fire_damage,
  MapNames = NULL, # use default
  LogFile = log_file,
  SummaryLogFile = sum_log_file
)

ext_dyn_fire$write()
```

### Dynamic Fuels

```{r dynamic-fuels-write}
## create the dynamic fuels extension config object
ext_dyn_fuel <- DynamicFuels$new(
  path = tmp_pth,
  Timestep = 10,
  SpeciesFuelCoefficients = spp_fuel_coeffs,
  HardwoodMaximum = 15L,
  DeadFirMaxAge = 15L, ## not needed w/o BDA extension
  FuelTypes = fuel_types,
  EcoregionTable = data.frame(FuelType = integer(0), Ecoregion = character(0)),
  DisturbanceConversionTable = disturb_conv,
  MapFileNames = NULL, ## use default
  PctConiferMapName = NULL, ## use default
  PctDeadFirMapName = NULL ## use default
)

ext_dyn_fuel$write()
```

## Verify configuration files

### Dynamic Fire

```{r dynamic-fire-config}
readLines(file.path(tmp_pth, ext_dyn_fire$files[1])) |>
  cat(sep = "\n")
```

*skip species table and raster files*

### Dynamic Fuels

```{r dynamic-fuels-config}
readLines(file.path(tmp_pth, ext_dyn_fuel$files[1])) |>
  cat(sep = "\n")
```

*skip species table and raster files*

## Cleanup

```{r cleanup}
withr::deferred_run()
```
