---
title: "Calibrating the Dynamic Fire System Extension"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating the Dynamic Fire System Extension}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
## All chunks are illustrative — they reference project-specific objects
## (e.g. `nfdb_point_fru59`) and spin up real LANDIS-II containers, so we
## disable both rendering (`eval = FALSE`) and code extraction
## (`purl = FALSE`) so `R CMD check` does not try to source them.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

```{r setup}
library(landisutils)
```

## Scope

This vignette covers the **calibration of the LANDIS-II Dynamic Fire System
extension** -- adjusting `SeverityCalibrationFactor`, the per-season FMC
`HiProp` values, and per-base-fuel-type `IgnProb` multipliers so simulated
fires match observed regional fire statistics.

It does **not** cover the empirical fits already done at the data layer
(`Mu`, `Sigma`, `Max`, `NumFires`, seasonal `PropFire`) -- those come from
NFDB lognormal MLE / count summaries and don't need an optimisation loop.

## Function families

The calibration is built from three layers:

| Layer | Purpose | Functions |
|---|---|---|
| Observation | Pre-compute observed targets from NFDB | `save_observed_fire_targets()` (+ `bc_fuel_code_to_base()` default; pass your own if fuel codes differ) |
| Scenario | Build a static-landscape calibration scenario (post-spinup IC; succession frozen; dynamic-fire baseline) | `build_calibration_spinup_scenario()`, `run_calibration_spinup()`, `build_calibration_scenario_template()` |
| Optimiser | Run DEoptim against the calibration scenario, applying patches per trial | `patch_fire_config()`, `loss_from_stats()`, `parse_dynamic_fire_logs()`, `sim_landis()` (or `sim_mock()` for testing), `calibrate_dynamic_fire()` |

Once calibrated, the best parameter vector flows back to a production fire
config via `apply_calibrated_ignprob()` + `apply_calibrated_hi_prop()` and
the `SeverityCalibrationFactor` scalar.

## Required project-side inputs

The library is project-agnostic but expects you to provide:

1. **NFDB-style SpatVectors** clipped to your primary fire regime (and
   optionally a secondary, broader regime for context) -- with `YEAR`
   and `SIZE_HA` columns on the points.
2. **A fuel-type raster** covering your LANDIS simulation extent, with a
   mapping from integer codes to base fuel types (`Conifer`,
   `ConiferPlantation`, `Deciduous`, `Slash`, `Open`). Use
   `bc_fuel_code_to_base()` if you're working with BC `FUEL_TYPE_CD`
   factor levels.
3. **A production scenario directory** with ForCS / Dynamic Fuels /
   Dynamic Fire configs already written. The calibration scenario template
   is built by copying this and patching.
4. **A spun-up IC**: produced by `build_calibration_spinup_scenario()` +
   `run_calibration_spinup()` -- one LANDIS-II run during pipeline build.

## End-to-end function sketch

The functions are designed to be composed inside a project's `targets`
pipeline (or any other workflow framework). Below is the typical wiring;
each numbered block is one or more `tar_target()` calls in practice.

```{r workflow-sketch}
## 1. Observed fire targets (NFDB-derived; cheap; no LANDIS-II).
save_observed_fire_targets(
  primary_points   = nfdb_point_fru59,
  primary_polys    = nfdb_poly_fru59,
  secondary_points = nfdb_point_frt12, ## optional regional context
  secondary_polys  = nfdb_poly_frt12,
  fire_years       = 1950L:2023L,
  fuel_types_rast  = fuel_types_rast,
  path             = "outputs/calibration/observed_fire_targets.rds",
  primary_label    = "FRU59",
  secondary_label  = "FRT12"
  ## fuel_code_to_base defaults to bc_fuel_code_to_base(); override if needed
)

## 2. Pre-calibration spinup: one LANDIS-II run with both ForCS spinup flags ON
##    to populate biomass; snapshot via the Output Biomass Community extension.
##    The snapshot's year-0 CSV+TIF become the calibration IC.
build_calibration_spinup_scenario(
  out_dir      = "LANDIS-II/_calibration_spinup",
  template_dir = "LANDIS-II/phase_2_ICH",          ## any non-fire scenario will do
  duration     = 1L,
  cell_length  = 100L
)
spinup_csv <- run_calibration_spinup(
  scenario_dir = "LANDIS-II/_calibration_spinup",
  base_seed    = 12345L,
  method       = "docker"
)
spinup_tif <- file.path(dirname(spinup_csv), "output-community-0.tif")

## 3. Calibration scenario template: copy from a production fire scenario, swap
##    in the spinup IC, patch ForCS to skip succession, write a fresh baseline
##    dynamic-fire.txt from the supplied tables.
build_calibration_scenario_template(
  out_dir                    = "LANDIS-II/_calibration",
  template_dir               = "LANDIS-II/phase_2_ICH_fire",
  snapshot_ic_csv            = spinup_csv,
  snapshot_ic_tif            = spinup_tif,
  baseline_fire_size_table   = fire_size_table,
  baseline_fuel_type_table   = defaultFuelTypeTable(),
  baseline_fire_damage_table = defaultFireDamageTable(),
  baseline_seasons_sim_table = seasons_sim_table,
  sim_years                  = 10L,
  cell_length                = 100L
)

## 4. DEoptim driver: starts a warm Docker pool, runs the calibration, tears
##    down on exit. Returns best_params (a named numeric of length 9) + the
##    DEoptim result object + a trace CSV.
cfg <- list(
  lower = c(
    SeverityCalibrationFactor = 0.5,
    SpHiProp = 0, SumHiProp = 0, FallHiProp = 0,
    IgnProb_Conifer = 0, IgnProb_ConiferPlantation = 0,
    IgnProb_Deciduous = 0, IgnProb_Slash = 0, IgnProb_Open = 0
  ),
  upper = c(
    SeverityCalibrationFactor = 2.5,
    SpHiProp = 1, SumHiProp = 1, FallHiProp = 1,
    IgnProb_Conifer = 1.5, IgnProb_ConiferPlantation = 1.5,
    IgnProb_Deciduous = 1.5, IgnProb_Slash = 1.5, IgnProb_Open = 1.5
  ),
  NP = 60L, itermax = 100L, n_reps = 5L, sim_years = 10L,
  weights = c(count = 1, size = 1, area_fuel = 0, severity = 0),
  simulator = "landis", method = "docker",
  n_cores = max(1L, parallel::detectCores() - 2L), parallel = TRUE,
  base_seed = 12345L
)
result <- calibrate_dynamic_fire(
  observed_targets_path = "outputs/calibration/observed_fire_targets.rds",
  scenario_template     = "LANDIS-II/_calibration/scenario.txt",
  cfg                   = cfg,
  out_dir               = "outputs/calibration"
)
print(result$best_params)

## 5. Apply calibrated params back to the production fire config (e.g., when
##    writing dynamic-fire.txt for the production scenarios).
calibrated <- result$best_params
production_fuel_type_table <- apply_calibrated_ignprob(
  defaultFuelTypeTable(), calibrated
)
production_fire_size_table <- apply_calibrated_hi_prop(
  fire_size_table, calibrated
)
production_severity_factor <- calibrated[["SeverityCalibrationFactor"]]
```

## Wiring in `targets`

The library doesn't ship a target factory (deliberate -- target naming
conventions vary across consuming projects). Adopt the helpers as plain
function calls inside `tar_target()`. Three things to watch for:

- **Cycle avoidance.** If your production `tar_target` that writes
  `dynamic-fire.txt` consumes `calibrated_fire_params` (the result of
  `calibrate_dynamic_fire()`), then the calibration's scenario template
  must NOT transitively depend on that target. Use
  `build_calibration_scenario_template()`'s `baseline_*` arguments to
  write a fresh uncalibrated dynamic-fire.txt inline, and depend on the
  per-extension *file* targets that populate your production scenario
  directory directly, NOT on your scenario-assembly target.

- **Deployment mode.** The calibration target should run on the main
  process (`deployment = "main"` in `targets`), not on a `crew` worker,
  because the function internally manages its own FORK cluster of
  `n_cores` workers + a warm Docker pool of `n_cores` containers.
  A crew worker trying to spawn its own cluster would compete for cores.

- **DEoptim is a Suggest.** Install via `renv::install("DEoptim")` before
  running the calibration target.

## Per-trial cost and lever knobs

`sim_landis()` per-trial cost is dominated by the LANDIS-II Docker run
(~1-3 minutes including dotnet startup). Total wall-clock for production
calibration: roughly `NP * itermax * n_reps * per-trial / n_cores`. The
warm Docker pool (started internally by `calibrate_dynamic_fire()`) saves
container-creation overhead but not dotnet startup, so the dominant cost
remains the LANDIS-II simulation itself. Practical levers if calibration
takes too long:

- Cut `sim_years` to 5 (instead of 10): roughly half the per-trial cost.
- Aggregate `fuel_types_rast` to a coarser resolution (e.g., 200 m cells)
  before passing to `build_calibration_scenario_template()`: ~4x speedup.
- Reduce `NP` × `itermax`; DEoptim's evolutionary search compensates
  somewhat for fewer evaluations.

A pure-R reimplementation (`sim_r_reimpl()`, currently a stub) would
remove the dotnet bottleneck entirely but introduces its own maintenance
burden tracking LANDIS-II's evolving algorithm. See the design notes in
`_tmp_dynamic_fire_calibration.md` of the gitanyow-partial-harvest
project.

## Testing without Docker

`sim_mock()` returns plausibly-shaped per-trial output without invoking
LANDIS-II, suitable for end-to-end testing of the DEoptim driver's
control flow:

```{r mock-test}
cfg$simulator <- "mock"
cfg$NP <- 6L
cfg$itermax <- 2L
cfg$parallel <- FALSE
calibrate_dynamic_fire(
  observed_targets_path = "outputs/calibration/observed_fire_targets.rds",
  scenario_template     = "LANDIS-II/_calibration/scenario.txt",
  cfg                   = cfg,
  out_dir               = tempdir()
)
```

The mock varies its output with `par_vec` so DEoptim sees a non-trivial
loss surface (good for catching driver-wiring regressions). It is NOT a
substitute for the real LANDIS-II simulator -- the parameter optima
under `sim_mock()` are not meaningful.

## See also

- `vignette("Dynamic-Fire-Fuels", package = "landisutils")` -- preparing
  the Dynamic Fire and Dynamic Fuels extension inputs in the first place.
- `?DynamicFire` -- the R6 class that backs the uncalibrated baseline
  config writer used inside `build_calibration_scenario_template()`.
- `?landis_pool_start` -- the warm Docker pool that
  `calibrate_dynamic_fire()` starts internally; also usable standalone.
