---
title: "Calibrating Species Growth Curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating Species Growth Curves}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Every succession extension needs per-species growth parameters, and they are
usually set by eye against whatever reference curves and plot data a project has.
This vignette describes how to run that calibration as a pipeline step instead:
sweep the parameters, score the resulting curves against the references, and
report a recommendation.

The toolkit is **succession-agnostic**. It works for any extension that keys
growth and mortality shape per species but maximum ANPP and maximum biomass per
(ecoregion, species) -- which covers ForC Succession and Biomass Succession
alike, and is precisely the asymmetry the design exploits.

## Why a whole sweep fits in one simulation

Because shape is a species attribute and level is an (ecoregion, species)
attribute, a sweep can be encoded as a landscape rather than as repeated runs:

* one **pseudo-species** per (species, growth shape, mortality shape), a clone of
  the real species differing only in those two fields;
* one **pseudo-ecoregion** per (max ANPP, max biomass) combination;
* one cell per (pseudo-species, pseudo-ecoregion) pair.

`growth_calibration_design()` builds that table. Hundreds of combinations then
run in a single simulation, at the cost of one landscape. This follows the
approach of the `Biomass_speciesFactorial` module.

```{r}
design <- growth_calibration_design(growth_params, grid = sweep_grid)
```

`growth_calibration_partition()` splits a design too large to hold in memory into
batches, always cutting on cell boundaries so a mixed cell's cohorts stay
together. `growth_structure_design()` crosses the sweep with a landscape's own
cohort structures when single-cohort cells are not representative enough.

## Three properties that keep the ranking honest

These are worth understanding before wiring anything, because each fixes a way
the obvious implementation goes wrong.

**Only the extension's own curve is parametric.** Plot observations are condensed
non-parametrically by `growth_bin_observations()`, which bins on stand age and
takes a quantile per bin. Fitting an empirical growth equation through the plot
cloud first -- as `Biomass_speciesParameters` does with Chapman-Richards,
Gompertz and logistic forms -- is right there, because those equations are the
ones that module's own succession implementation uses. It is wrong anywhere else:
it inserts a curve family the extension does not use, and then matches the
extension's output to that intermediary rather than to the data.

**Every reference series carries the same weight.** `growth_reference_curves()`
evaluates all series on one common age grid, so a modelled curve contributing
several hundred points and a plot cloud contributing a dozen count equally.
Pooling raw observations instead makes the effective objective an accident of how
many rows each series happens to have -- in practice the species that score worst
are simply the ones with no modelled reference curve.

**Level is derived, not searched.** A cohort never quite reaches its
maximum-biomass parameter: it approaches the asymptote while mortality is already
removing biomass. The fraction it does achieve depends only on the shapes and on
the ratio of maximum ANPP to maximum biomass, not on the absolute level. So hold
maximum biomass fixed across the sweep, rank on shape alone, and recover the
level arithmetically with `growth_inflation_factor()`. Sweeping level alongside
shape lets the two trade off against each other, and the ranking then settles
wherever the reference data happen to be centred rather than where the growth
curve belongs.

Verify that invariance on your own extension before relying on it: run
combinations that share an ANPP-to-biomass ratio but differ in absolute maximum
biomass, and confirm they agree on the achieved fraction.

## Mapping your extension onto the canonical names

The toolkit uses four canonical column names. Rename your extension's parameters
onto them before calling, and rename back when promoting:

| Canonical | ForC Succession | Biomass Succession |
|---|---|---|
| `growth_shp` | `GrowthCurve` | `GrowthCurve` |
| `mort_shp` | `MortalityShape` | `MortalityShape` |
| `anpp_max` | `ANPPTimeSeries` max ANPP | `maxANPP` |
| `biomass_max` | `MaxBiomassTimeSeries` | `maxBiomass` |

The one genuinely extension-specific quantity is `biomass_max_scale`: divide the
maximum-biomass parameter by it to express that parameter in whatever units the
curves are in.

```{r}
## ForC Succession: biomass_max is g/m^2 of biomass, the summary log reports
## g C/m^2, and curves are conventionally Mg C/ha -- so 200.
growth_score_fit(curve, ref, biomass_max_scale = 200)

## Biomass Succession: maxBiomass and the log are both g/m^2 of biomass.
growth_score_fit(curve, ref, biomass_max_scale = 1)
```

Get this wrong and the shape ranking is unaffected -- it is scale-free -- but
every recovered `biomass_max_est` is off by the ratio of the two scales. Check it
against a species whose level you already trust.

## What stays on the project side

Deliberately not in this package, because each is project- or
extension-specific:

* **the succession config writer** -- the calibration landscape needs a
  species table, ecoregions, initial communities, a climate file and a scenario
  file, and the succession extension's own config;
* **the output reader** -- extensions write different logs, and the column
  holding aboveground biomass differs;
* **reference assembly** -- which curves and plot datasets exist, and how they
  are read, varies by project. Produce a data frame with `source`, `age` and a
  value column, and the toolkit takes it from there;
* **the parameter table itself** -- see the promotion loop below.

## Run it as a separate targets project

Give the calibration its own `{targets}` project rather than folding it into an
existing pipeline. `{targets}` takes an exclusive lock on the store for the whole
of `tar_make()`, so a shared store makes it impossible to calibrate while another
phase of the same pipeline is running. Separate stores remove the contention, and
the calibration then needs none of the main pipeline's inputs.

Hand results across by **file**, not by cross-store `tar_read()`: a report cannot
track a target in another store as a dependency, and would stop rendering from a
clean checkout.

A minimal chain:

```{r}
list(
  ## references -> windows -> curves, computed once per species
  tar_target(windows, growth_fitting_windows(references, species_core, scoring)),
  tar_target(
    ref_curves,
    growth_reference_curves(
      references[[sp]],
      window = growth_window_for(windows, sp),
      bin = ctl$age_bin,
      plot_quantile = ctl$plot_quantile
    ),
    pattern = map(sp), iteration = "list"
  ),

  ## sweep -> landscape -> one run -> curves
  tar_target(design, growth_calibration_design(growth_params, grid = grid)),
  tar_target(inputs, write_my_calibration_inputs(design, ...), format = "file"),
  tar_landis(name = run, scenario_dir = dir, deps = list(inputs)),
  tar_target(curves, read_my_calibration_curves(run, design)),

  ## score -> rank -> report
  tar_target(scores, {
    curves |>
      dplyr::group_split(map_code) |>
      purrr::map(\(cur) dplyr::bind_cols(
        meta(cur),
        growth_score_fit(cur, ref_curves[[species_of(cur)]])
      )) |>
      dplyr::bind_rows() |>
      growth_add_objective()
  }),
  tar_target(best, growth_best_candidates(scores, growth_params, windows, scoring)),
  tar_target(review, write_growth_review_bundle(dir, sp, curves, ...), format = "file")
)
```

Targets that write paths tracked by version control should be
`deployment = "main"`: a worker writing into its own checkout dirties it, which
breaks fast-forward-only node synchronisation.

## The fitting window is derived, not nominated

Nobody should have to pick an age range by hand. `growth_auto_window()` opens the
window at an age floor below which ground-plot programmes do not sample, and
closes it at the earliest of a quantile of observed plot ages, the end of the
reference curve, and a fraction of `longevity`.

That last bound is the one that binds in practice. LANDIS-II ramps mortality up
as a cohort approaches `longevity` and the curve then falls to exactly zero, so
an open-ended window scores the modelled die-off rather than the level the stand
holds. The cap is a fraction of `longevity` rather than something read off the
simulated curve, because decline timing depends on the mortality shape, which is
itself being swept -- a candidate-dependent window would score different
candidates over different ranges and could not rank them fairly.

Bounds come back as whole years, since the model steps annually.

## Put the judgement calls in a file

Several choices in this workflow are judgements, not facts, and they should be
recorded where a reviewer can see and change them rather than buried in code.
`read_growth_scoring()` reads a per-species table:

| Column | Meaning |
|---|---|
| `age_min`, `age_max` | Explicit window bounds. Blank keeps the derived window. |
| `age_bin` | Width of the age bins the plot cloud is condensed into. |
| `plot_quantile` | Quantile within each bin. `0.5` tracks central tendency; higher moves toward the upper envelope. |
| `plots_warn_below` | Advisory. Below this the fit is flagged sparse, not refused. |
| `weight_*` | Relative weight of each reference series in the ranking. `0` drops a series from the score while leaving it on the figures. |
| `level_source` | Which reference's plateau the recommended maximum biomass follows. |

The weights matter most. Modelled yield curves describe **potential** growth for
fully stocked, pure, undisturbed stands -- which is exactly what a single-cohort
calibration cell is. Ground plots are **realised** stands spanning every stocking
level and disturbance history, usually with whole-plot volume attributed to a
leading species that holds well under all of it. Which a species should follow is
a decision worth recording rather than letting it fall out of whichever reference
happened to have more rows.

Two behaviours exist to stop the pipeline inventing answers:

* a species with no scorable reference comes back `fitted = FALSE` with no
  parameters, rather than a ranking over indistinguishable rows returning
  whichever sorted first;
* a nominated `level_source` is a **constraint**. If that reference has no level,
  no recommendation is made, instead of silently substituting another one.

## Promotion is manual, and should stay that way

The calibration reports; it never writes the parameter table. The loop is:

1. edit the sweep grid and the scoring table;
2. run the calibration project;
3. review the per-species figures and the summary, treating the argmin as a
   diagnostic -- a parameter whose sensitivity boxes overlap is one the data
   cannot constrain, and copying its "best" value is fitting noise;
4. hand-edit the parameter table for the values you accept, and record what
   justified the change;
5. re-run, and check that downstream simulations are actually invalidated. A
   value can change, flow into the generated configs, and still not reach the
   outputs if runs are pinned. This is the step that is easy to get wrong.

`write_growth_review_bundle()` writes step 3's artifacts to a plain directory
needing no report render: per-species figures with the fitting window shaded and
the binned plot series drawn over the references, plus a summary table of
promotable parameters beside the values in use.
